Performance comparison of Dask and Apache Spark on HPC systems \HL{for Neuroimaging}

### 2022-09-12 latex/paper.tex
small fixes

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\newcommand{\HL}[1]{\hl{#1}}

@@

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly used
	in the field, it is important to evaluate performance of different engines.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has substantial performance benefits in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems,
	it is hard to reserved multiple nodes concurrently on those infrastructure.
	Moreover, by using our infrastructure, we can assure that no external processes
	are influencing the benchmark for our experiements.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and 
	replicates as close as possible its configuration.
	This lets us compare Spark and Dask in an isolated environement with the condition 
	of a realistic  allocation of resources, for neuroimaging applications, in an HPC system.
}

@@

\HL{
	We chose these dataset as they represent common dataset size for neuroimaging 
	imaging studies and for their free access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects from various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application.
}

@@

\HL{
	To minimize the performance difference from implementation decisions, the 
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the engine scheduling and
	their APIs.
}

The scripts used for our experiments and the results generated are available on
GitHub at: \mbox{\href{https://github.com/mathdugre/paper-big-data-engines}{https://github.com/mathdugre/paper-big-data-engines/}}.

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this applications for its real-life use case, large data shuffling between workers,
and requirements to keep all data in-memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and it is containerized unlike our applications.}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	With a large amount of worker threads accessing the Lustre file system
	concurrently, the disk bandiwdth or network are saturated and it create 
	an I/O bottleneck to access the Lustre file system.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently they risk 
	of saturation the shared disk bandwidth, thus resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	it reduces the contention on the disk bandwidth resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what is the cause for reporting extra Spark processes.
	However, with our analysis we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, the is no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\HL{
	In this study, we observed that Dask has a larger startup overhead that Spark.
	These results are consistent with a previous study~{\cite{Mehta:17}}.
	However, other studies found the opposite~{\cite{8588652}} or that the difference was negligible~{\cite{8943502}}
}
Those apparent contractions most likely come from differences in infrastructure and configuration of the engines.

@@

\subsection{Recommendations}
\iffalse
The differences in performance are minimal between both engines. However, 
we added a “Recommendation” subsection to the discussion for qualitative 
difference and some caveats to avoid when using the engines.
\fi
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark is minimal, in general. 
	The main differences we observed are qualitative.
	We found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging application.
	Spark configuration required to set multiple environment variables.
	In comparison, Dask is ready-to-go after an installation through PIP.
	Dask dashboard provides more information than Spark ones, making
	the debugging of Dask application easier and faster.
	Therefore, for the neuroimaging field, we recommend the use of Dask over Spark.
	However, in other fields with non-Python-based ecosystem, both Dask and spark
	would be equal candidate as a Big Data engine.
}

>>>>

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\newcommand{\HL}[1]{\hl{#1}}
\newcommand{\TG}[1]{\color{red}\textsc{From Tristan: }#1\color{black}}

@@

\section{Introduction}
The rise in data sharing coupled with improved data collection technologies leads neuroimaging into the Big Data era\cite{ALFAROALMAGRO2018400, van2014human, ConpPortal}.
While standard neuroimaging workflow engines, such as Nipype\cite{Nipype:11}, are well rounded to process standard compute-intensive pipelines,
they lack Big Data strategies (i.e., in-memory computing, data locality, and lazy-evaluation) to improve the performance of increasingly prevalent data-intensive pipelines.
A previous study\cite{8752675} noted the importance of in-memory computing and data locality to improve the performance of data-intensive pipelines.
This work studies how performance benchmarks \HL{of neuroimaging applications} generalize across Big Data engines in an HPC environment.
\HL{
	With the rising need for Big Data engines in neuroimaging and none commonly used
	in the field, it is important to evaluate performance of different engines.
}
In particular, we study in-depth the performance difference between Dask\cite{Dask:15} and Apache Spark\cite{Spark:16} for their suitability to process neuroimaging pipelines.
\HL{
	Our main contribution is to assess whether Spark or Dask has any substantial performance advantage in processing neuroimaging pipelines in HPC systems.
}

@@

\HL{
	While our infrastructure is not of the scale of TOP500 HPC systems,
	we can assure that no external processes
	impacted our measurements which would be hard to achieve on a production system.
	Our system uses similar hardware as Compute Canada's B\'eluga cluster
	--- a representative cluster available to researchers in Canada --- and 
	replicates as closely as possible its configuration.
	This lets us compare Spark and Dask in an isolated environement with the condition 
	of a realistic  allocation of resources, for neuroimaging applications, in an HPC system.
}

@@

\HL{
	We chose these dataset as they represent common dataset sizes for neuroimaging 
	imaging studies and are open-access.
}

\subsection{Applications}
This section describes the applications used to benchmark Dask and Spark.
We use three simple synthetic applications to have a deep understanding of the 
underlying behavior of the applications with simple I/O patterns, namely
\textit{Increment}, \textit{Multi-Increment}, and \textit{Histogram}.
Also, two neuroimaging applications are used to study more realistic 
applications, namely \textit{Kmeans} and \textit{BIDS App Example}{\cite{gorgolewski2017bids}}.
\HL{
	We chose these neuroimaging data processing applications to study the 
	effects from various Big Data characteristics.
	Table~{\ref{table:applications}} depicts the Big Data characteristics used in
	each application. 
}
\TG{Explain what each column represents}

@@

\HL{
	To minimize the performance difference from implementation decisions, the 
	code base of the applications is common for both engines.
	Therefore, the performance observed comes directly from the engine scheduling and
	their APIs.
} The scripts used for our experiments and the results generated are available on
GitHub at: \mbox{\href{https://github.com/mathdugre/paper-big-data-engines}{https://github.com/mathdugre/paper-big-data-engines/}}.

@@

\subsubsection{Kmeans}
For our third application, we apply Kmeans clustering to the voxel
intensities of the BigBrain image. We set the number of clusters to 3. The
application starts by reading the image blocks, combining all voxels in a
1-D array, and choosing initial centroids using the min, max, and
intermediate values. It assigns each voxel to its closest centroid and
updates each centroid with the average of its assigned voxels. It repeats
the assignment and updates steps for a configurable number of iterations.
Finally, the voxels of the image blocks are classified and written back to
the file system. Updating the centroids involves substantial data
communication between the workers.
The computation to find the closest centroids for each voxels scales with the number of centroids.
It can be computed in parallel at the cost of higher memory or sequentially to reduce memory usage.
We decided to compute the nearest centroids in parallel since we use only three centroids.
However, due to resources limitation on our infrastructure, the Kmeans experiments
are performed with only half of the BigBrain dataset ({\SI{303}{\gibi\byte}}).
\HL{We chose this application as it involves substantial data shuffling between workers,
and requires to keep all the data in memory.}

\subsubsection{BIDS App Example}
Our fourth application is the BIDS App Example, a neuroimaging pipeline to
measure brain volume from MRIs. For this application, we use the CoRR
dataset. The application extracts the brain volume of each participant
then computes the average for each group of participants. Unlike the other
applications, the BIDS App Example is a command-line executed in a Docker image
(bids/example on DockerHub). We converted the Docker image to a Singularity
image for use in HPC environments, using
\href{https://hub.docker.com/r/singularityware/docker2singularity/tags/}{docker2singularity}
\HL{We chose this application for its real-life use case and the fact that it is containerized unlike our other applications.}

@@

\section{Results}
\subsection{Increment}
Figure \ref{fig:increment_worker} depicts the total execution time spent in each code segment of the \textit{Increment} application.
The bars are broken down into Overhead, Read, Write, and other application-specific functions.
The mean of each bar is calculated from ten repetitions, with the error bar representing the standard deviation.
As expected, the compute time, from \textit{Increment}, remains the same when we vary the number of nodes.
However, the \textit{Read} and \textit{Write} time increases with the number of nodes for both engines.
\HL{
	Due to the large amount of worker threads accessing the Lustre file system
	concurrently, the disk bandiwdth or network are saturated, which results in an I/O bottleneck.
}

From Figure \ref{fig:increment_worker}, we also observe that Dask's overhead is higher than Spark's.
While the overhead of both engines increases with the number of nodes, Dask's overhead is significantly worse when scaling.
Although Dask has a higher overhead, the total time for both engines is overall similar.
This is due to a tradeoff between disk bandwidth and compute/overhead time.
\HL{
	On the one hand, when multiple threads access Lustre concurrently they may saturate the shared disk bandwidth, resulting in longer I/O time.
	On the other hand, when most threads are either idle or performing computations,
	the contention on the disk bandwidth is reduced, resulting in shorter I/O time.
}
The only significant difference in total time is when two nodes are used.
This results in an approximately 3\% total runtime difference.
\begin{figure}[!h]
	\centering
	\includegraphics[clip,width=0.75\columnwidth]{figures/stacked_increment_worker.jpg}
	\caption{Increment: varying nodes -- 5000 files, 5 iterations}
	\label{fig:increment_worker}
\end{figure}

@@

In a follow-up experiment, we further optimized our Kmeans implementations for both Dask and Spark to minimize the size of data kept in memory.
Figure~{\ref{fig:dask_kmeans_gantt}} and Figure~{\ref{fig:spark_kmeans_gantt}} show the Gantt chart for the execution
of this new experiments for Dask and Spark.
The x-axis shows the time in second while the y-axis depicts the workers, with 
one worker thread ID (Dask) and process ID (Spark) per line.
We represent each task on the Gantt chart with its associated function's color.
We observe that Spark reports three times more processes than Dask threads.
\HL{
	It is unclear what is the cause for reporting extra Spark processes.
	However, we found that both Dask and Spark use the same 
	amount of maximum concurrent processes or threads.
	Additionally, there was no significant difference in the average amount of 
	concurrent processes or threads between Dask and Spark.
}

@@

\HL{
	In this study, we observed that Dask has a larger startup overhead that Spark.
	These results are consistent with a previous study~{\cite{Mehta:17}}.
	However, other studies found the opposite~{\cite{8588652}} or that the difference was negligible~{\cite{8943502}}.
}
Those apparent contractions most likely come from differences in infrastructure and configuration of the engines.

@@

\subsection{Recommendations}
\iffalse
The differences in performance are minimal between both engines. However, 
we added a “Recommendation” subsection to the discussion for qualitative 
difference and some caveats to avoid when using the engines.
\fi
\HL{
	The results from our experiments show that the performance difference between
	Dask and Spark is minimal, in general. 
	The main differences we observed are qualitative.
	We found that Dask integration with the scientific Python ecosystem made it
	easier to implement neuroimaging applications.
	Spark configuration required to set multiple environment variables.
	In comparison, Dask is ready-to-go after an installation through PIP, the Python package manager.
	The Dask dashboard provides more information than Spark's, making
	the debugging of Dask applications easier and faster.
	Therefore, for the neuroimaging field, we recommend the use of Dask over Spark.
	However, in other fields with non-Python-based ecosystem, both Dask and Spark
	may be equal candidates as a Big Data engine.
}
